Thermal reactionomes reveal divergent responses to thermal extremes in warm

### 2015-06-03 doc/ApTranscriptome_manuscript.Rmd
revisions for MolEco

John Stanton-Geddes

Department of Biology, University of Vermont, 109 Carrigan Dr, Burlington, VT 05405.

@@

Email: john.stantongeddes.research@uvm.edu

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced defense*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with *enhanced defense* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.3°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) and characterized the reaction norm for each gene using a custom analysis. We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adaptation.

@@

Ideally, genetically-based variation in gene expression profiles would be identified by comparing individuals completely reared under common-garden conditions to eliminate environmental variation experienced either as adults or during development. However, *Aphaenogaster* colonies are long-lived, cannot be bred under laboratory conditions, and do not achieve complete turnover of the workforce for at least a year or longer. Thus, as is commonly done with other long-lived organisms [@barshis2013, ], we exposed both colonies to common-garden rearing conditions for three months to fully acclimate adult workers to common temperatures. Over this time, roughly 1-2 cohorts of new  workers are expected to join the workforce (~1/3 of the total), such that the workers sampled for thermal traits and gene expression are likely to have included a mix of adult-acclimated and fully lab-reared individuals.

To characterize the thermal reactionome, we haphazardly collected three ants from each 2012 colony at the same time on 12 days. Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. We extracted mRNA by homogenizing the three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Methods). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.

@@

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. As preliminary examination of the data (Supplemental Methods) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental Methods).

@@

We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental Methods).

For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°C) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.

@@

The two species differed in the number of transcripts in each response category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`). *Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A. carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`). Consistent with the *enhanced defense* hypothesis, *A. picea* induced 1,146 (30%) more transcripts in response to low temperatures than *A. carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`). However, consistent with the predictions of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, at high temperatures *A. carolinensis* induced expression of ~14% fewer transcripts than *A. picea* (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* < `r max(chi5$p.value, 0.001)`).

@@

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another. These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Fig. S1) indicating that these differences are not simply due to neutral processes. The *enhanced defense* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species. The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species. For each temperature extreme, we tested these predictions by focusing on the genes upregulated by the less-adapted species (*A. carolinensis* for low temperatures, *A. picea* for high temperatures) and testing whether those genes displayed the same response profile in the better-adapted species. As with the overall results, the patterns of association differed for the two temperature extremes. Genes upregulated at low temperatures in *A. carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from all other expression categories in *A. picea*, suggesting shared response pathways (Fig. 2A). In contrast, genes upregulated in response to high temperatures in *A. picea* (**High** and **Bimodal**) were only weakly associated with this same category in *A. carolinensis* (Fig. 2B). Instead, these genes were less likely than expected to be upregulated in any context, with a significant excess in the **Not Responsive** and **Intermediate** categories (Fig. 2B).

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A. carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 3A). For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 3A).

@@

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 3A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* may induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.

@@

Given the differences in the patterns of thermal responsiveness between species (Fig. 2, 2), it is worth noting a number of similarities. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Table 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role. Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig. 3). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.

Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009]. While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in , expression type categories or critical temperature thresholds as we do here (Fig. 2, Fig. S1). Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression, as opposed to identifying lists of differentially-expressed genes, can provide meaningful insight on the process of evolution.

As with all *de novo* transcriptomes, our work has some caveats. First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness. In particular, functional protein levels cannot be expected to be directly linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability (...citations…).

@@

>>>>

John Stanton-Geddes, Data Scientist

Dealer.com, 1 Howard St, Burlington, VT 05401

@@

Email: john.stantongeddes.research@google.com

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with *enhanced response* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. While temporal patterns of transcriptional activity (e.g. fast vs. slow responding genes) are also important components of an organism's transcriptional response to environmental conditions [@sorensen2005], we focus of the response of transcripts across conditions at the same time point. We use the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.3°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we measured the reaction norm for each gene using a regression-based statistical approach to identify temperature-dependent patterns of change in gene expression. We used these response patterns to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced response hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways [@neelakanta2012; @franssen2011]. Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adaptation.

@@

Ideally, genetically-based variation in gene expression profiles would be identified by comparing individuals completely reared under common-garden conditions to eliminate environmental variation experienced either as adults or during development. However, *Aphaenogaster* colonies are long-lived, cannot be bred under laboratory conditions, and do not achieve complete turnover of the workforce for at least a year or longer. Thus, as is commonly done with other long-lived organisms [@whitehead2006; @barshis2013], we exposed both colonies to common-garden rearing conditions for six months to fully acclimate adult workers to common temperatures. Over this time, roughly 1-2 cohorts of new workers are expected to join the workforce (~1/3 of the total), such that the workers sampled for thermal traits and gene expression are likely to have included a mix of adult-acclimated and fully lab-reared individuals.

Unlike ANOVA-based experimental designs, which derive statistical power from replication within each experimental treatment level, regression designs have greater power when sampling additional values across the range of the continuous predictor variable [@cottingham2005]. Thus, we focused our sequencing efforts on maximizing the number of temperatures at which the transcriptome was profiled, rather than on replication at each temperature. To limit differences in gene expression not related to the experimental treatment, we haphazardly collected three ants from each 2012 colony at the same time on 12 days. We measured response to temperature with a one-hour static temperature application, which is ecologically-relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012]. Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum (~0°C) and maximum (~43°C) temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. Thus, our reactionome characterized early, but not late, responding genes. We extracted mRNA by homogenizing the three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supporting Information). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.

@@

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. As preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).

@@

@@

The two species differed in the number of transcripts in each response category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`). *Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A. carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`). Consistent with the *enhanced response* hypothesis, *A. picea* induced 1,146 (30%) more transcripts in response to low temperatures than *A. carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`). However, consistent with the predictions of the *tolerance hypothesis* but not the *enhanced response hypothesis*, at high temperatures *A. carolinensis* induced expression of ~14% fewer transcripts than *A. picea* (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* < `r max(chi5$p.value, 0.001)`).

@@

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another. These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Fig. S1) indicating that these differences are not simply due to neutral processes. The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species. The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species. For each temperature extreme, we tested these predictions by focusing on the genes upregulated by the less-adapted species (*A. carolinensis* for low temperatures, *A. picea* for high temperatures) and testing whether those genes displayed the same response profile in the better-adapted species. As with the overall results, the patterns of association differed for the two temperature extremes. Genes upregulated at low temperatures in *A. carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from all other expression categories in *A. picea*, suggesting shared response pathways (Fig. 2A). In contrast, genes upregulated in response to high temperatures in *A. picea* (**High** and **Bimodal**) were only weakly associated with this same category in *A. carolinensis* (Fig. 2B). Instead, these genes were less likely than expected to be upregulated in any context, with a significant excess in the **Not Responsive** and **Intermediate** categories (Fig. 2B).

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *tolerance hypothesis* but not the *enhanced response hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A. carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 3A). For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 3A).

@@

Under the *enhanced response* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 3A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* may induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.

@@

Given the differences in the patterns of thermal responsiveness between species (Fig. 2), it is worth noting a number of similarities. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures (Table 1). The degree of upregulation at low temperatures is surprising given previous studies [@teets2012-1; @vesala2012] that found little transcriptional activity at low temperatures. However, these studies exposed organisms to a few extreme (-10 -- 0°C) temperatures. At these extreme low temperatures we also found few upregulated transcripts (Fig. 3A), while the peak of low-temperature transcriptional activity occurred near 10°C (Fig. 3). A potential explanation for this pattern is that increased gene expression is required to support elevated metabolism at cold temperatures, as suggested by the metabolic cold adaptation hypothesis [@addo-bediako2002]. The observation that more transcripts were upregulated at low than high temperatures could be due to stronger selection on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role. Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig. 3). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

A major contribution of this study is the construction of a reactionome for gene expression data (see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach). For example, the degree of upregulation at cool (~10°C) but not extreme cold temperatures was missed in previous studies that focused on extreme cold limits, as we discussed above. Further, a number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.

Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009]. While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in expression type categories or critical temperature thresholds as we do here (Fig. 2, Fig. S1). Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression, as opposed to identifying lists of differentially-expressed genes, can provide meaningful insight on the process of evolution.

A number of caveats do apply to our work. First, species may differ in gene expression along axes which we have not measure here, primarily temporal patterns of gene expression [@sorensen2005], and could be studied in futher work. Second, the *de novo* transcriptome assembly is highly fragmented, given that all sequenced ant genomes to date have only about 18,000 genes [@gadau2012]. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Third, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness. In particular, functional protein levels cannot be expected to be directly linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability [@feder2005].

@@

![Fig. 1](../results/predictions_Fig1.png)
